---
title: "ML Model Comparison for Fish Threat in R"
author: "Elke Windschitl"
date: "2022-12-09"
format: html
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```

Description: In this qmd, I evaluate different supervised machine learning algorithms for predicting IUCN Red List status of fish based on ecological and morphological characteristics. These characteristics were retrieved from FishBase and joined with the IUCN data in a separate script.

## Introduction

Global human activity threatens many species with extinction. According to the International Union and Conservation of Nature (IUCN), “More than 41,000 species are threatened with extinction. That is still 28% of all assessed species.” [1]. Increased extinction and loss of biodiversity can have severe ecological, economic, and cultural impacts. Cardinale et al.’s deep dive into biodiversity and ecosystem services research conclude that biodiversity loss reduces ecological communities’ efficiency, stability, and productivity. Decreased productivity from ecosystem services can have a negative impact on ecosystem economics [2]. Additionally, cultures worldwide have strong ties to local flora and fauna, much of which now face extinction risk. Improving understanding of extinction risk is ecologically, economically, and culturally important.

Wildlife scientists have been working to understand what ecological traits of vertebrates predict threat level, and what common risk factors drive those threat level rates. Munstermann et al. investigate what terrestrial vertebrate functional groups are most at risk of extinction threat and find that cave dwelling amphibian, arboreal quadrupedal mammals, aerial and scavenging birds, and pedal squamates are at high risk [3]. This knowledge can help inform policies and practices with the goal to decrease threats of extinction of wildlife. However, less comprehensive research has been done to conduct similar analyses on marine species.

The IUCN Red List has many species that are listed as "Data Deficient" or "Not Evaluated". Filling in these data gaps is extremely important when it comes to conservation. In marine species, evaluating these populations can prove challenging. It can be helpful to build off of existing knowledge to inform where evaluation resources should be spent. Here, I propose to build a machine learning model that predicts binary Red List status of saltwater fish based on their ecological and morphological traits according to FishBase. I then apply the most successful model to Red List Data Deficient and Not Evaluated species. 

This work builds off of my previous work [Identifying Key Traits in Hawaiian Fish that Predict Risk of Extinction](https://elkewind.github.io/posts/2022-12-02-hawaiian-fish-analysis/). However, here I am looking at all fish listed on the IUCN Red List -- not just those in Hawaii -- and I am using a Tidymodels machine learning approach.

## The Data

For my analyses I use the IUCN Red List data accessed via the IUCN Red List API [1] and package rredlist [4]. Consistent with Munstermann et al., living species listed as ‘Vulnerable’, ‘Endangered’, or ‘Critically Endangered’ were categorized as ‘Threatened’. Living species listed as ‘Least Concern’ and ‘Near Threatened’ were categorized as ‘Nonthreatened’ [3]. The IUCN Red List data are limited in that many marine species have not been listed yet or have been identified as too data deficient to be evaluated. The lack of data on elusive fish may introduce bias into the models.

Fish ecological data were accessed from FishBase [5] via package rfishbase [6]. Different species in the FishBase data were originally described by different people, possibly leading to errors or biases. Measurement errors in length may be present, as there are various common ways to measure the length of a fish. The species recorded in FishBase may be biased towards fish with commercial value. Data were wrangled in R and formatted in a tidy data table with the following variables.

```{r include=FALSE}
library(tidyverse)
library(knitr)
# Read in data
fish_dat <- read_csv("/Users/elkewindschitl/Documents/data-sci/fish_data.csv") %>%
  filter(!is.na(IsOfConcern)) %>% 
  slice_head(n = 10) %>% 
  kable()
```

## Methods

To get started, there are several packages I will be using. *Tidyverse* packages help with further cleaning and preparing data. *Tidymodels* packages have almost all of what I need for the machine learning steps. *kknn* helps me build my knn model. *hrbrthemes* and *viridis* are used to quickly make aesthetically pleasing looking figures. *knitr* is used to create kable tables. *baguette* is used in my bagging model. *doParallel* allows for parallel computing on my laptop. *vip* helps to identify variable importance.

```{r}
# Load libraries
library(tidyverse)
library(tidymodels)
#library(rsample)   
#library(recipes)
#library(skimr)
library(kknn)
library(hrbrthemes)
library(viridis)
library(knitr)
#library(workflows)
library(baguette)
library(doParallel)
#library(forcats)
#library(glmnet)
library(vip)
```

First I read in the data. These data were cleaned and joined in a separate script, but they will still need a bit of preprocessing. The outcome variable in this dataset is labeled IsOfConcern and indicates if the species is at risk or extinction (1) or not (0). I start out by exploring the data dimensions. 

```{r}
# Read in data
fish_dat_full <- read_csv("/Users/elkewindschitl/Documents/data-sci/fish_data.csv") 
fish_dat <- fish_dat_full %>%
  filter(!is.na(IsOfConcern)) # remove columns that don't have outcome variable

# Explore some characteristics of the dataset
cols <- ncol(fish_dat)
rows <- nrow(fish_dat)
df_chars <- data.frame(
  Metric = c("Number of Columns", "Number of Rows"),
  Count = c(ncol(fish_dat), nrow(fish_dat)))
kable(df_chars,
      col.names = c("", "Count"))

fish_dat %>% 
  group_by(IsOfConcern) %>%
  count() %>% 
  kable(col.names = c("Species treat is of concern", "Count"))
```

### Data prep

There are a lot on NA values in this dataset. I have a lot of columns already, so I can reduce that by removing columns that have a high proportion of NA values. Here I only keep columns where less than 20% of rows have NA values.

```{r}
# Calculate the proportion of NA values in each column
na_proportion <- colMeans(is.na(fish_dat), na.rm = TRUE)

#I want to remove rows with extreme NA counts (more than 20%)
# Define the threshold (20% or 0.20)
threshold <- 0.20

# Find columns with more than the threshold proportion of NA values
columns_meeting_threshold <- names(na_proportion[na_proportion <= threshold])

# Print the column names that meet the threshold
print(columns_meeting_threshold) %>% kable(col.names = "Columns that are below NA threshold")

# Select for just those columns that meet my criteria
fish_short <- fish_dat %>% 
  select(all_of(columns_meeting_threshold))
```

There is still more to be done. I want to make sure numeric columns are numeric, and that character columns are treated as factors. I need to make sure my outcome variable is a factor as well. I have a small enough data frame that I am able to do this by looking at which columns are character and mutating them to be factors. I want unordered factors, and I neeed to remove columns that are identifiers rather than features. Because some of the algorithms I am working with here do not handle missing data well, I chose to remove all of the rows that had NA values. This did unfortunately cut down on the amount of data that I have to train and test the algorithms on.

```{r}
# Find character columns that need to be converted to factor
sapply(fish_short, class) %>% kable(col.names = c("Column", "Class"))

# List of character columns to convert to factors
character_columns_to_convert <- c("GenusSpecies", "BodyShapeI", "DemersPelag", "AirBreathing", "PriceCateg", "UsedforAquaculture", "Dangerous", "Electrogenic", "MainCommonName")

# Convert the specified character columns to factors
fish <- fish_short %>%
  mutate(across(all_of(character_columns_to_convert), as.factor))

# If feature is a factor DON'T order, remove identifying columns
fish <- fish %>% mutate_if(is.ordered, .funs = factor, ordered = F)  %>% 
  select(-GenusSpecies) %>% 
  select(-SpecCode) %>% 
  select(-MainCommonName)

# Make outcome factor
fish$IsOfConcern <- as.factor(fish$IsOfConcern)

# Remove rows with any remaining missing values (shouldn't be any, but in case)
fish <- na.omit(fish)

# Check the new df
sapply(fish, class) %>% kable(col.names = c("Column", "Class"))
```

After my data are prepped, I need to split the data into training and testing data sets. I use a 70/30 split. I have unbalanced data, so I stratafy by my outcome variable, IsOfConcern

```{r}
set.seed(123)
# Initial split of data, default 70/30
fish_split <- initial_split(fish, prop = 0.7, strata = IsOfConcern)
fish_train <- training(fish_split)  # Training data
fish_test <- testing(fish_split)    # Test data

```

### Preprocessing

I create a recipe for the preprocessing steps used. I use dummy columns to make all the factor (categorical) variables have their own column. I remove columns where there is no variation in the data. Then I normalize the numeric columns because the lasso and knn algorithms require normalization to avoid certain features dominating the model. I use the same preprocessing steps for all algorithms for adequate comparison. 

```{r}
# Preprocess the data within the recipe
fish_recipe <- recipe(IsOfConcern ~ ., data = fish_train) %>% 
  step_dummy(all_factor(), -all_outcomes(), one_hot = TRUE) %>% 
  step_zv(all_predictors()) %>%  
  step_normalize(all_numeric(), -all_outcomes())

# Check test and train dfs look as expected
prepped <- fish_recipe %>% 
  prep()
fish_baked_train <- bake(prepped, fish_train)
fish_baked_test <- bake(prepped, fish_test)

# Use below to check for NA values in the entire dataframes
#any(is.na(fish_baked_train))
#any(is.na(fish_baked_test))
```

### Dummy Classifier

Because my data are unbalanced with many more non-threatened species, if a model always chose non-threatened it would have a high accuracy. Of course, that is not very helpful when trying to predict with species that *might be* threatened. Here I derive a dummy accuracy by calculating the accuracy of a model that always predicts non-threatened. This will serve as a baseline for if a model is performing well (better than the dummy) or not. However, because this dataset is so imbalanced, I will be using area under the curve (AUC) for model selection.

```{r}
# Calculate dummy classifier for baseline comparison
# Calculate the number of rows where IsOfConcern is 0
num_is_0 <- sum(fish_test$IsOfConcern == 0)

# Calculate the number of rows where IsOfConcern is not 0
num_is_not_0 <- nrow(fish_test) - num_is_0

# Calculate the accuracy of the dummy classifier (always predicting the majority class)
dummy <- num_is_0 / nrow(fish_test)
```

The dummy classifier accuracy is `{r} dummy`. This will serve as the baseline for other algorithms.

### Lasso for Classification

```{r}
set.seed(123)

# Set up k-fold cross validation with 10 folds. This can be used for all the algorithms
fish_cv = fish_train %>% 
  vfold_cv(v = 10,
           strata = IsOfConcern)

# Set specifications
tune_l_spec <- logistic_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet")

# Define a workflow
wf_l <- workflow() %>%
  add_model(tune_l_spec) %>% 
  add_recipe(fish_recipe)
# set grid
lambda_grid <- grid_regular(penalty(), levels = 50)

doParallel::registerDoParallel()
set.seed(123)

# Tune lasso model
lasso_grid <-   wf_l %>% 
  tune_grid(
    add_model(tune_l_spec),
    resamples = fish_cv,
    grid = lambda_grid
)

# Plot the mean accuracy and AUC at each penalty
lasso_grid %>%
  collect_metrics() %>%
  ggplot(aes(penalty, mean, color = .metric)) +
  geom_errorbar(aes(ymin = mean - std_err,
                    ymax = mean + std_err),
                alpha = 0.5) +
  geom_line(size = 1.5) +
  facet_wrap(~.metric, 
             scales = "free", 
             strip.position = "left",
             nrow = 2, labeller = as_labeller(c(`accuracy` = "Accuracy", 
                                                `roc_auc` = "Area under ROC curve"))) +
  scale_x_log10(name = "Penalty") +
  scale_y_continuous(name = "") +
  scale_color_manual(values = c("#4a6c75", "#57ba72")) +
  theme_minimal() +
  theme(
    strip.placement = "outside",
    legend.position = "none",
    panel.background = element_blank(),
    plot.background = element_blank()
  )

# View table
lasso_grid %>%
  tune::show_best(metric = "roc_auc") %>%
  slice_head(n = 5) %>% 
  kable(caption = "Performance of the best models and the associated estimates for the penalty parameter values.")

# Select the model with the highest auc
best_lasso <- lasso_grid %>%
  select_best("roc_auc")

final_l_wf <- wf_l %>% 
  finalize_workflow(best_lasso)

# Perform a last fit to see how the model performs on the test data
final_lasso_fit <- last_fit(final_l_wf, fish_split)

# Collect metrics on the test data
tibble_lasso <- final_lasso_fit %>% collect_metrics()
tibble_lasso %>% 
  kable(caption = "Accuracy and area under ther receiver operator curve of the final fit.")

# Grab the model accuracy on the testing data
final_lasso_accuracy <- tibble_lasso %>%
  filter(.metric == "accuracy") %>%
  pull(.estimate)
final_lasso_auc <- tibble_lasso %>%
  filter(.metric == "roc_auc") %>%
  pull(.estimate)

# Bind predictions and original data
lasso_test_rs <- cbind(fish_test, final_lasso_fit$.predictions)[, -16]# Remove duplicate column

# Compute a confusion matrix
cm_lasso <- lasso_test_rs %>% yardstick::conf_mat(truth = IsOfConcern, estimate = .pred_class) 

# Create a custom color palette
custom_palette <- scale_fill_gradient(
  high = "#4a6c75",   
  low = "#d3e6eb"  
)

# Create the confusion matrix heatmap plot
autoplot(cm_lasso, type = "heatmap") +
  custom_palette +  # Apply the custom color palette
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14),
    panel.background = element_blank(),
    plot.background = element_blank()
  ) +
  labs(title = "Confusion matrix of lasso predictions on test data.")

# Calculate rates of tru pos, false neg. etc. from the confusion matrix
TP_las <- cm_lasso$table[2, 2]
FP_las <- cm_lasso$table[2, 1]
TN_las <- cm_lasso$table[1, 1]
FN_las <- cm_lasso$table[1, 2]
TPR_las <- TP_las / (TP_las + FN_las)  # True Positive Rate
FPR_las <- FP_las / (FP_las + TN_las)  # False Positive Rate
TNR_las <- TN_las / (TN_las + FP_las)  # True Negative Rate
FNR_las <- FN_las / (TP_las + FN_las)  # False Negative Rate

# Create cm df to hold all false pos, etc. metrics
lasso_cm_vec <- c(TPR_las, FPR_las, TNR_las, FNR_las)
row_names <- c("True positive rate", "False positive rate", "True negative rate", "False negative rate")
cm_df <- bind_cols(Metric = row_names, Lasso = lasso_cm_vec)
```

The accuracy for the lasso model was `r final_lasso_accuracy` which is slightly better than our dummy classifier that had an accuracy of `r dummy`. This model had an AUC of `r final_lasso_auc`. 

### K-Nearest Neighbor

```{r}
# Define the KNN model with tuning
knn_spec_tune <- nearest_neighbor(neighbors = tune()) %>% # tune k
  set_mode("classification") %>% 
  set_engine("kknn")

# Define a new workflow
wf_knn_tune <- workflow() %>% 
  add_model(knn_spec_tune) %>% 
  add_recipe(fish_recipe)
    
# Fit the workflow on the predefined folds and hyperparameters
fit_knn_cv <- wf_knn_tune %>% 
  tune_grid( 
    fish_cv, 
    grid = data.frame(neighbors = c(1,5,10,15,seq(20,200,10))))

# Use autoplot() to examine how different parameter configurations relate to accuracy
autoplot(fit_knn_cv) +
  theme_light() +
  labs(
    x = "Number of neighbors (K)"
  ) +
  theme(
    legend.position = "none",
    panel.background = element_blank(),
    plot.background = element_blank()
  ) +
  facet_wrap(
    ~.metric,
    nrow = 2,
    labeller = labeller(.metric = c("accuracy" = "Accuracy", "roc_auc" = "Area under ROC curve"))
  )

# View table
fit_knn_cv %>%
  tune::show_best(metric = "roc_auc") %>%
  slice_head(n = 5) %>% 
  kable(caption = "Performance of the best models and the associated estimates for the number of neighbors parameter values.")

# Select the model with the highest auc
best_knn <- fit_knn_cv %>%
  select_best("roc_auc")

# The final workflow for our KNN model
final_knn_wf <-
  wf_knn_tune %>% 
  finalize_workflow(best_knn)

# Use last_fit() approach to apply model to test data
final_knn_fit <- last_fit(final_knn_wf, fish_split)

# Collect metrics on the test data
tibble_knn <- final_knn_fit %>% collect_metrics()
tibble_knn %>% 
  kable(caption = "Accuracy and area under ther receiver operator curve of the final fit.")

# Store accuracy and AUC
final_knn_accuracy <- tibble_knn %>%
  filter(.metric == "accuracy") %>%
  pull(.estimate)
final_knn_auc <- tibble_knn %>%
  filter(.metric == "roc_auc") %>%
  pull(.estimate)

# Bind predictions and original data
knn_test_rs <- cbind(fish_test, final_knn_fit$.predictions)[, -16]

# Compute a confusion matrix
cm_knn <- knn_test_rs %>% yardstick::conf_mat(truth = IsOfConcern, estimate = .pred_class) 

# Create the confusion matrix heatmap plot
autoplot(cm_knn, type = "heatmap") +
  custom_palette +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14),
    panel.background = element_blank(),
    plot.background = element_blank()
  ) +
  labs(title = "Confusion matrix of knn predictions on test data.")


# Calculate rates from the confusion matrix
TP_knn <- cm_knn$table[2, 2]
FP_knn <- cm_knn$table[2, 1]
TN_knn <- cm_knn$table[1, 1]
FN_knn <- cm_knn$table[1, 2]
TPR_knn <- TP_knn / (TP_knn + FN_knn)  # True Positive Rate
FPR_knn <- FP_knn / (FP_knn + TN_knn)  # False Positive Rate
TNR_knn <- TN_knn / (TN_knn + FP_knn)  # True Negative Rate
FNR_knn <- FN_knn / (TP_knn + FN_knn)  # False Negative Rate

# Add rates to cm df
knn_cm_vec <- c(TPR_knn, FPR_knn, TNR_knn, FNR_knn)
cm_df$KNN <- knn_cm_vec
```

We see here that our k-nearest neighbors model had nearly the same accuracy at predicting threat status than the dummy classifier. The accuracy of the model was `r final_knn_accuracy`. This model had an AUC of `r final_knn_auc` which is better than the lasso model. 

### Decision Tree

```{r}
# Tell the model that we are tuning hyperparams
tree_spec_tune <- decision_tree(
  cost_complexity = tune(),
  tree_depth = tune(),
  min_n = tune()) %>% 
  set_engine("rpart") %>% 
  set_mode("classification")

# Set up grid
tree_grid <- grid_regular(cost_complexity(), tree_depth(), min_n(), levels = 5)

# Check grid
#tree_grid 

# Define a workflow with the recipe and specification
wf_tree_tune <- workflow() %>% 
  add_recipe(fish_recipe) %>% 
  add_model(tree_spec_tune)

doParallel::registerDoParallel(cores = 3) #build trees in parallel

# Tune
tree_rs <- tune_grid(
  wf_tree_tune,
  IsOfConcern~.,
  resamples = fish_cv,
  grid = tree_grid,
  metrics = metric_set(roc_auc)
)

# Use autoplot() to examine how different parameter configurations relate to auc
autoplot(tree_rs) + 
  theme_light() +
  scale_color_manual(values = c("#4a6c75", "#57ba72", "#d596e0", "#e06d53", "#d6cf81")) + 
  labs(x = "Cost-complexity parameter",
       y = "Area under the ROC curve") +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    panel.background = element_blank(),
    plot.background = element_blank()
  )

# View table
tree_rs %>%
  tune::show_best(metric = "roc_auc") %>%
  slice_head(n = 5) %>% 
  kable(caption = "Performance of the best models and the associated estimates for the tuned tree parameter values.")

# Finalize the model specs with the best hyperparameter result
final_tree <- finalize_model(tree_spec_tune, select_best(tree_rs))

# Final fit to test data 
final_tree_fit <- last_fit(final_tree, IsOfConcern~., fish_split) # does training fit then final prediction as well

# Collect metrics from fit
tibble_tree <- final_tree_fit %>% collect_metrics()
tibble_tree %>% kable(caption = "Accuracy and area under ther receiver operator curve of the final fit.")

# Store accuracy and auc metrics
final_tree_accuracy <- tibble_tree %>%
  filter(.metric == "accuracy") %>%
  pull(.estimate)

final_tree_auc <- tibble_tree %>%
  filter(.metric == "roc_auc") %>%
  pull(.estimate)

# Bind predictions and original data
tree_test_rs <- cbind(fish_test, final_tree_fit$.predictions)[, -16]

# Compute a confusion matrix
cm_tree <- tree_test_rs %>% yardstick::conf_mat(truth = IsOfConcern, estimate = .pred_class) 

# Create the confusion matrix heatmap plot
autoplot(cm_tree, type = "heatmap") +
  custom_palette +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14),
    panel.background = element_blank(),
    plot.background = element_blank()
  ) +
  labs(title = "Confusion matrix of decision tree predictions on test data.")


# Calculate rates from the confusion matrix
TP_tree <- cm_tree$table[2, 2]
FP_tree <- cm_tree$table[2, 1]
TN_tree <- cm_tree$table[1, 1]
FN_tree <- cm_tree$table[1, 2]
TPR_tree <- TP_tree / (TP_tree + FN_tree)  # True Positive Rate
FPR_tree <- FP_tree / (FP_tree + TN_tree)  # False Positive Rate
TNR_tree <- TN_tree / (TN_tree + FP_tree)  # True Negative Rate
FNR_tree <- FN_tree / (TP_tree + FN_tree)  # False Negative Rate

# Add rates to cm df
tree_cm_vec <- c(TPR_tree, FPR_tree, TNR_tree, FNR_tree)
cm_df$DecisionTree <- tree_cm_vec
```

We see here that our decision tree model had a higher accuracy at predicting threat status than the dummy classifier or other models. The accuracy of the decision tree was `r round(final_tree_accuracy, 3)`. The AUC is `r round(final_tree_auc, 3)` which is higher than the lasso model and similar to the knn model.

### Bagging

```{r}
set.seed(123)
# Set bagging tuning specifications
bag_spec <- 
  bag_tree(cost_complexity = tune(),
  tree_depth = tune(),
  min_n = tune()) %>% 
  set_engine("rpart", times = 75) %>% # 25 ensemble members 
  set_mode("classification")

# Set up tuning grid
bag_grid <- grid_regular(cost_complexity(), tree_depth(), min_n(), levels = 5)

# Check the grid space
#bag_grid

# Set up bagging workflow
wf_bag <- workflow() %>% 
  add_recipe(fish_recipe) %>% 
  add_model(bag_spec)

doParallel::registerDoParallel() #build trees in parallel

# Run tuning
bag_rs <- tune_grid(
  wf_bag,
  IsOfConcern~.,
  resamples = fish_cv,
  grid = bag_grid,
  metrics = metric_set(roc_auc)
)

# Use autoplot() to examine how different parameter configurations relate to auc
autoplot(bag_rs) + 
  theme_light() +
  scale_color_manual(values = c("#4a6c75", "#57ba72", "#d596e0", "#e06d53", "#d6cf81")) + 
  labs(x = "Cost-complexity parameter",
       y = "Area under the ROC curve") +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    panel.background = element_blank(),
    plot.background = element_blank()
  )

# View table
bag_rs %>%
  tune::show_best() %>%
  slice_head(n = 5) %>% 
  kable(caption = "Performance of the best models and the associated estimates for the tuned tree parameter values.")

# Finalize the model specs witht he best performing model
final_bag <- finalize_model(bag_spec, select_best(bag_rs))

# Perform a last fit of the model on the testing data
final_bag_fit <- last_fit(final_bag, IsOfConcern~., fish_split)

# Collect metrics from fit
tibble_bag <- final_bag_fit %>% collect_metrics()
tibble_bag %>% 
  kable(caption = "Accuracy and area under ther receiver operator curve of the final fit.")

# Store model accuracy and auc on testing data
final_bag_accuracy <- tibble_bag %>%
  filter(.metric == "accuracy") %>%
  pull(.estimate)
final_bag_auc <- tibble_bag %>%
  filter(.metric == "roc_auc") %>%
  pull(.estimate)

# Bind predictions and original data
bag_test_rs <- cbind(fish_test, final_bag_fit$.predictions)[, -16]

# Compute a confusion matrix
cm_bag <- bag_test_rs %>% yardstick::conf_mat(truth = IsOfConcern, estimate = .pred_class) 

# Create the confusion matrix heatmap plot
autoplot(cm_bag, type = "heatmap") +
  custom_palette +  # Apply the custom color palette
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14),
    panel.background = element_blank(),
    plot.background = element_blank()
  ) +
  labs(title = "Confusion matrix of bagging predictions on test data.")


# Calculate rates from the confusion matrix
TP_bag <- cm_bag$table[2, 2]
FP_bag <- cm_bag$table[2, 1]
TN_bag <- cm_bag$table[1, 1]
FN_bag <- cm_bag$table[1, 2]
TPR_bag <- TP_bag / (TP_bag + FN_bag)  # True Positive Rate
FPR_bag <- FP_bag / (FP_bag + TN_bag)  # False Positive Rate
TNR_bag <- TN_bag / (TN_bag + FP_bag)  # True Negative Rate
FNR_bag <- FN_bag / (TP_bag + FN_bag)  # False Negative Rate

# Add rates to cm df
bag_cm_vec <- c(TPR_bag, FPR_bag, TNR_bag, FNR_bag)
cm_df$Bagging <- bag_cm_vec
```

We see here that our bagging model had a *lower* accuracy at predicting threat status than the decision tree and a similar accuracy as the dummy classifier. The accuracy of the bagging was `r round(final_bag_accuracy, 3)`. The AUC is `r round(final_bag_auc, 3)`.

### Random Forest

```{r}
set.seed(123)

# Set forest specifications
forest_spec <- 
  rand_forest(min_n = tune(),
              mtry = tune(),
              trees = tune()) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# Set grid for tuning
forest_grid <- grid_regular(min_n(), mtry(), trees(), levels = 5)

# Check the grid space
#forest_grid

# Set the workflow with the recipe and forest specs
wf_forest <- workflow() %>% 
  add_recipe(fish_recipe) %>% 
  add_model(forest_spec)

doParallel::registerDoParallel()

# Perform tuning
forest_rs <- tune_grid(
  wf_forest,
  IsOfConcern~.,
  resamples = fish_cv,
  grid = forest_grid,
  metrics = metric_set(roc_auc)
)

# Use autoplot() to examine how different parameter configurations relate to auc
autoplot(forest_rs) + 
  theme_light() +
  scale_color_manual(values = c("#4a6c75", "#57ba72", "#d596e0", "#e06d53", "#d6cf81")) + 
  labs(x = "# Randomly selected predictors",
       y = "Area under the ROC curve") +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    panel.background = element_blank(),
    plot.background = element_blank()
  )

# View table
forest_rs %>%
  tune::show_best() %>%
  slice_head(n = 5) %>% 
  kable(caption = "Performance of the best models and the associated estimates for the tree and forest parameter values.")

# Finalize model with specs and best hyperparameters
final_forest <- finalize_model(forest_spec, select_best(forest_rs))

# Perform last fit
final_forest_fit <- last_fit(final_forest, IsOfConcern~., fish_split)

# Collect performance metrics
tibble_forest <- final_forest_fit %>% collect_metrics()
tibble_forest %>% 
  kable(caption = "Accuracy and area under ther receiver operator curve of the final fit.")

# Store accuracy and auc metrics
final_forest_accuracy <- tibble_forest %>%
  filter(.metric == "accuracy") %>%
  pull(.estimate)
final_forest_auc <- tibble_forest %>%
  filter(.metric == "roc_auc") %>%
  pull(.estimate)

# Bind predictions and original data
forest_test_rs <- cbind(fish_test, final_forest_fit$.predictions)[, -16]

# Compute a confusion matrix
cm_forest <- forest_test_rs %>% yardstick::conf_mat(truth = IsOfConcern, estimate = .pred_class) 

# Create the confusion matrix heatmap plot
autoplot(cm_forest, type = "heatmap") +
  custom_palette +  # Apply the custom color palette
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14),
    panel.background = element_blank(),
    plot.background = element_blank()
  ) +
  labs(title = "Confusion matrix of random forest predictions on test data.")


# Calculate rates from the confusion matrix
TP_forest <- cm_forest$table[2, 2]
FP_forest <- cm_forest$table[2, 1]
TN_forest <- cm_forest$table[1, 1]
FN_forest <- cm_forest$table[1, 2]
TPR_forest <- TP_forest / (TP_forest + FN_forest)  # True Positive Rate
FPR_forest <- FP_forest / (FP_forest + TN_forest)  # False Positive Rate
TNR_forest <- TN_forest / (TN_forest + FP_forest)  # True Negative Rate
FNR_forest <- FN_forest / (TP_forest + FN_forest)  # False Negative Rate

# Add rates to cm df
forest_cm_vec <- c(TPR_forest, FPR_forest, TNR_forest, FNR_forest)
cm_df$RandomForest <- forest_cm_vec
```

The accuracy of the forest was `r round(final_forest_accuracy, 3)`. This is slightly higher than the dummy classifier, but still lower that the decision tree. This model had an auc of `r round(final_forest_auc, 3)`.

### Boosting

```{r}
# Tune learning rate first
# Set up specs for learning rate tuning
lr_spec <- parsnip::boost_tree(mode = "classification",
                                engine = "xgboost",
                                trees = 3000,
                                learn_rate = tune())

# Set up tuning grid
lr_grid <- expand.grid(learn_rate = seq(0.0001, 0.5, length.out = 50))

# Set up workflow
wf_lr_tune <- workflow() %>% 
  add_recipe(fish_recipe) %>% 
  add_model(lr_spec)

doParallel::registerDoParallel()
set.seed(123)

# Tune
lr_rs <- tune_grid(
  wf_lr_tune,
  IsOfConcern~.,
  resamples = fish_cv,
  grid = lr_grid
)

# Use autoplot() to examine how different parameter configurations relate to accuracy 
autoplot(lr_rs) +   
  theme_light() +
  labs(
    x = "Learning rate"
  ) +
  theme(
    legend.position = "none",
    panel.background = element_blank(),
    plot.background = element_blank()
  ) +
  facet_wrap(
    ~.metric,
    nrow = 2,
    labeller = labeller(.metric = c("accuracy" = "Accuracy", "roc_auc" = "Area under ROC curve"))
  )

# Identify best values from the tuning process
lr_rs %>%
  tune::show_best(metric = "roc_auc") %>%
  slice_head(n = 5) %>% 
  kable(caption = "Performance of the best models and the associated estimates for the learning rate parameter values.")

# Select best lr hyperparametes
best_learn <- lr_rs %>%
  tune::select_best("roc_auc")

# Tune tree parameters next
# Create a new specification where setting the learning rate and tune the tree parameters
boost_tree_spec <- parsnip::boost_tree(mode = "classification",
                                engine = "xgboost",
                                trees = 3000,
                                learn_rate = best_learn$learn_rate,
                                min_n = tune(),
                                tree_depth = tune(),
                                loss_reduction = tune()
                                )

# Define parameters to be tuned
boost_params <- dials::parameters(
  min_n(),
  tree_depth(),
  loss_reduction()
)

# Set up a tuning grid using grid_max_entropy() to get a representative sampling of the parameter space.
boost_tree_grid <- dials::grid_max_entropy(boost_params, size = 50)

# Set up workflow
wf_boost_tree_tune <- workflow() %>% 
  add_recipe(fish_recipe) %>% 
  add_model(boost_tree_spec)

set.seed(123)
doParallel::registerDoParallel()

# Tune
boost_tree_rs <- tune_grid(
  wf_boost_tree_tune,
  IsOfConcern~.,
  resamples = fish_cv,
  grid = boost_tree_grid
)

# Identify best values from the tuning process
boost_tree_rs %>%
  tune::show_best(metric = "roc_auc") %>% 
  slice_head(n = 5) %>% 
  kable(caption = "Performance of the best models and the associated estimates for the tree parameter values.")

# Select best tree hyperparameters
boost_best_trees <- boost_tree_rs %>%
  tune::select_best("roc_auc")

# Tune Stochastic Parameters
# Create another new specification where setting the learning rate and tree parameters and tune the stochastic parameters.
boost_stoc_spec <- parsnip::boost_tree(mode = "classification",
                                engine = "xgboost",
                                trees = 3000,
                                learn_rate = best_learn$learn_rate,
                                min_n = boost_best_trees$min_n,
                                tree_depth = boost_best_trees$tree_depth,
                                mtry = tune(),                   
                                loss_reduction = boost_best_trees$loss_reduction,
                                sample_size = tune(),
                                stop_iter = tune()
                                )

# Set up a tuning grid using grid_max_entropy() again.
# Define parameters to be tuned
boost_stoc_params <- dials::parameters(
  finalize(mtry(),
           select(fish_baked_train,-IsOfConcern)),
  sample_size = sample_prop(c(.4, .9)),
  stop_iter())

# Set up grid
boost_stoc_grid <- dials::grid_max_entropy(boost_stoc_params, size = 50)

# Set up workflow
wf_boost_stoc <- workflow() %>% 
  add_recipe(fish_recipe) %>% 
  add_model(boost_stoc_spec)

set.seed(123)
doParallel::registerDoParallel()

# Tune
boost_stoc_rs <- tune_grid(
  wf_boost_stoc,
  IsOfConcern~.,
  resamples = fish_cv,
  grid = boost_stoc_grid
)

# Identify best values from the tuning process
boost_stoc_rs %>%
  tune::show_best(metric = "roc_auc") %>%
  slice_head(n = 5) %>% 
  kable(caption = "Performance of the best models and the associated estimates for the stochastic parameter values.")

# Select best hyperparameters from tuning
boost_best_stoch <- boost_stoc_rs %>%
  tune::select_best("roc_auc")

# Finalize workflow
# Assemble final workflow with all of the optimized parameters and do a final fit.
boost_final_spec <- parsnip::boost_tree(mode = "classification",
                                engine = "xgboost",
                                trees = 1000,
                                learn_rate = best_learn$learn_rate,
                                min_n = boost_best_trees$min_n,
                                tree_depth = boost_best_trees$tree_depth,
                                mtry = boost_best_stoch$mtry,                   
                                loss_reduction = boost_best_trees$loss_reduction,
                                stop_iter = boost_best_stoch$stop_iter,
                                sample_size = boost_best_stoch$sample_size
                                )

# Set up workflow
wf_boost_final <- workflow() %>% 
  add_recipe(fish_recipe) %>% 
  add_model(boost_final_spec)

# Fit to just training data (need for later)
final_simple_fit <- wf_boost_final %>%
  fit(data = fish_train)

# Final fit
final_boost_fit <- last_fit(boost_final_spec, IsOfConcern~., fish_split)

# Store accuracy and auc metrics
tibble_boost <- final_boost_fit %>% collect_metrics()
tibble_boost %>% 
  kable(caption = "Accuracy and area under ther receiver operator curve of the final fit.")

final_boost_accuracy <- tibble_boost %>%
  filter(.metric == "accuracy") %>%
  pull(.estimate)

final_boost_auc <- tibble_boost %>%
  filter(.metric == "roc_auc") %>%
  pull(.estimate)

# Bind predictions and original data
boost_test_rs <- cbind(fish_test, final_boost_fit$.predictions)[, -16]

# Compute a confusion matrix
cm_boost <- boost_test_rs %>% yardstick::conf_mat(truth = IsOfConcern, estimate = .pred_class) 

# Create the confusion matrix heatmap plot
autoplot(cm_boost, type = "heatmap") +
  custom_palette +  # Apply the custom color palette
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14),
    panel.background = element_blank(),
    plot.background = element_blank()
  ) +
  labs(title = "Confusion matrix of random boosted predictions on test data.")


# Calculate rates from the confusion matrix
TP_boost <- cm_boost$table[2, 2]
FP_boost <- cm_boost$table[2, 1]
TN_boost <- cm_boost$table[1, 1]
FN_boost <- cm_boost$table[1, 2]
TPR_boost <- TP_boost / (TP_boost + FN_boost)  # True Positive Rate
FPR_boost <- FP_boost / (FP_boost + TN_boost)  # False Positive Rate
TNR_boost <- TN_boost / (TN_boost + FP_boost)  # True Negative Rate
FNR_boost <- FN_boost / (TP_boost + FN_boost)  # False Negative Rate

# Add rates to cm df
boost_cm_vec <- c(TPR_boost, FPR_boost, TNR_boost, FNR_boost)
cm_df$Boosting <- boost_cm_vec
```

The accuracy of the boosting was `r round(final_boost_accuracy, 3)`. This is higher than the dummy classifier, but slightly lower that the decision tree. This model had an auc of `r round(final_forest_auc, 3)`, which is the best of all models.

## Results

### Model selection

I want to compare accuracy and area under the curve of all models created. 

```{r}
# Name models in vec
models <- c("Dummy", "Lasso", "KNN", "Decision Tree", "Bagging", "Random Forest", "Boosting")

# Create accuracy vec
accuracy <- c(dummy, final_lasso_accuracy, final_knn_accuracy, final_tree_accuracy, final_bag_accuracy, final_forest_accuracy, final_boost_accuracy)

# Make df
accuracy_df <- data.frame(models, accuracy)

# Create a factor with the desired order for models
accuracy_df$models <- factor(accuracy_df$models, levels = c("Dummy", "Lasso", "KNN", "Decision Tree", "Bagging", "Random Forest", "Boosting"))

# Create the plot
ggplot(accuracy_df, aes(x = models, y = accuracy)) +
  geom_col(fill = "#4a6c75") +
  theme_minimal() +
  labs(title = "Accuracy was similar across all models",
    x = NULL,        
    y = "Accuracy") +
  theme(plot.background = element_blank(),
        panel.background = element_blank())

# Create auc vec
auc <- c(final_lasso_auc, final_knn_auc, final_tree_auc, final_bag_auc, final_forest_auc, final_boost_auc)

# Make df
auc_df <- data.frame(models[-1], auc)
print(auc_df)

# Create a factor with the desired order for models
auc_df$models <- factor(auc_df$models, levels = c("Lasso", "KNN", "Decision Tree", "Bagging", "Random Forest", "Boosting"))

ggplot(auc_df, aes(x = models, y = auc)) +
  geom_col(fill = "#4a6c75") +
  theme_minimal() +
  labs(title = "Boosting was the highest performing model",
    x = NULL,        
    y = "Area under ROC curve") +
  theme(plot.background = element_blank(),
        panel.background = element_blank())
```

Because the dataset is very imbalanced with the proportion of threatened species being small, the accuracies between models don't vary much. It is difficult to get much more accurate than always choosing not threatened (the dummy classifier) because that already has an accuracy of `r dummy`. All models performed better than the dummy classifier though, so that is encouraging. Looking at AUC we see that boosting decision trees was the most successful of the models when applied to the testing data. 

### Varibale Importance

I want to look at the variable importance of the most successful model, the boosting.

```{r}
# Variable importance
var_imp_boost <- wf_boost_final %>% 
  fit(fish_train) %>% 
  pull_workflow_fit() %>%
  vi() %>%
  mutate(
    Importance = abs(Importance),
    Variable = fct_reorder(Variable, Importance))
var_imp_boost %>% 
  ggplot(aes(x = Importance, y = Variable)) +
  geom_col(fill = "#57ba72") +
  theme_minimal() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = "Feature",
       x = "Importance",
       title = "Feature importance of boosting model")
```

It looks like length, having no special electrogenic ability, other body shape, and depth range (deep) are driving the model the most. 